MWA tied-array processing III: Microsecond time resolution via a polyphase synthesis filter

### 2020-05-09 ipfb.tex
Chipping away at the new Quantisation Error section.

One of the great advantages of the critically sampled PFB is the existence of a mathematically equivalent but computationally efficient implementation.
It can be shown that Eq. \eqref{eqn:wdft} is equivalent to first segmenting the windowed time series into taps, summing their respective samples element-wise, and performing a single DFT on the resulting array (now also of size $K$), i.e.
\begin{equation}
    X_k[m] = \sum_{n=0}^{K-1} b[n]\,e^{-2\pi jkn/K},
    \label{eqn:pfb}
\end{equation}
where
\begin{equation*}
    b[n] = \sum_{\rho=0}^{P-1} h[n + K\rho]\,x[n + mM + K\rho].
\end{equation*}
A short proof of this equivalence is given in \citet{Harris2011}.
The procedure described by Eq. \eqref{eqn:pfb} is called the \textit{weighted overlap-add} algorithm, and is illustrated in Fig. \ref{fig:pfb}.
\begin{figure}[p]
    \centering
    (a) \\
    \includegraphics[width=\columnwidth]{cartoon-top.eps} \\
    (b) \\
    \includegraphics[width=\columnwidth]{cartoon.eps}
    \caption{A diagrammatic representation of the weighted overlap-add algorithm, as defined in Eq. \eqref{eqn:pfb}. Panel (a) shows the filter window being translated along a discretely sampled signal (in this case, containing a sinusoid and noise) with a step size of one tap. At each step, panel (b) shows how the signal (first row) is multiplied by a filter (second and third rows), and each tap is summed (bottom left) and Fourier transformed to produce the final spectrum (bottom right).}
    \label{fig:pfb}
\end{figure}

@@

\subsection{MWA implementation of the fine PFB}

@@

In the context of an astrophysical signal, it is desirable to quantify the reconstruction error in terms of the signal-to-noise ratio (S/N).
If we assume that individual samples follow, and are dominated by, the same Gaussian noise statistics, then the reduction in S/N of a reconstructed sample can be estimated by considering the fraction of the power of the reconstructed sample that came from the original sample:
\begin{equation}
    \frac{[\text{S/N}]_\text{recon}}{[\text{S/N}]_\text{orig}}
      = \frac{\left(\sum\limits_{m = -\infty}^{\infty} f_m[n]\,h_m[-n]\right)^2}{
            \sum\limits_{s = -\infty}^{\infty}
            \left(\sum\limits_{m = -\infty}^{\infty} f_m[n]\,h_{m-s}[-n]\right)^2
        }.
\end{equation}
This expression is invariant under the substitution $n \rightarrow n + \lambda M$, $\lambda \in \mathbb{Z}$, which indicates that the (average) reconstruction error is only a function of where the sample in question falls within a tap.
Consequently, any synthesis filter (except the exact inverse of the analysis filter, if it exists) will introduce a ``ringing'' effect into the reconstructed time series that has a period equal to the size of the PFB tap.

@@

Minimising the loss of S/N is equivalent to solving Eq. \eqref{eqn:invertibility_condition} in a least squares sense.
Since any given reconstructed sample only receives contributions from samples spaced one tap apart, Eq. \eqref{eqn:invertibility_condition} can be thought of as $M$ independent conditions, one for each tap position, $n = 0, 1, 2, \dots, M-1$.
Thus, considering each tap position separately, each condition can be expressed as a minimal matrix equation,
\begin{equation}
    H^{(n)}F^{(n)} = D,
\end{equation}
where $H^{(n)}$, $F^{(n)}$, and $D$ are matrices whose elements are given by
\begin{equation}
    \begin{aligned}
        H^{(n)}_{ij} &= h_{P-1+j-i}[-n], \\
        F^{(n)}_j    &= f_j[n], \\
        D_i          &= \delta^{P-1}_i,
    \end{aligned}
\end{equation}
where $P$ is the number of taps in the analysis filter, and the size of $F^{(n)}$ is set to the desired number of (non-zero) taps in the synthesis filter, $P^\prime$.
The indices are chosen such that, owing to the finite size of the analysis filter, the smallest $H^{(n)}$ that captures every non-trivial term in Eq. \eqref{eqn:invertibility_condition} is the $(P^\prime+P-1) \times P^\prime$ matrix
\begin{equation}
    H^{(n)} =
    \begin{bmatrix}
       h_{P-1}[-n] & 0 & \cdots & 0 \\
       h_{P-2}[-n] & h_{P-1}[-n] & \cdots & 0 \\
       \vdots & & \ddots & \vdots \\
       h_0[-n] & h_1[-n] & \cdots & h_{P-1}[-n] \\
       \vdots & & \ddots & \vdots \\
       0 & \cdots & h_0[-n] & h_1[-n] \\
       0 & \cdots & 0 & h_0[-n]
    \end{bmatrix},
\end{equation}
with row number $i = P - 1$ (in $H^{(n)}$ and $D$) corresponding to the $s = 0$ term.

Once the matrices $H^{(n)}$, $F^{(n)}$, and $D$ have been defined, solution by least squares regression can proceed in the usual way, yielding the best-fit filter coefficients
\begin{equation}
    \hat{F}^{(n)} = \left({H^{(n)}}^T H^{(n)}\right)^{-1} {H^{(n)}}^T D,
\end{equation}
where ${H^{(n)}}^T$ is the transpose of $H^{(n)}$.
With this notation, the reconstruction error can be more simply expressed
\begin{equation}
    \frac{[\text{S/N}]_\text{recon}}{[\text{S/N}]_\text{orig}}
      = \frac{(\hat{D}_{P-1})^2}{\hat{D}^T \hat{D}},
    \label{eqn:snr_short}
\end{equation}
where $\hat{D} = H^{(n)} \hat{F}^{(n)}$.

Fig. \ref{fig:synthfilters} shows the solutions found for the MWA's analysis filter defined in Eq. \eqref{eqn:filter}, for 12-, 18-, and 24-tap synthesis filter sizes.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{example-image-a}
    \caption{Least-squares solutions for different filter sizes \dots}
    \label{fig:synthfilters}
\end{figure}
Despite the complexity of the synthesis filters, they all share the same basic form as the analysis filter.
This suggests that choosing a synthesis filter that is the mirror image of the analysis filter (i.e. $f[n] = h[-n]$), might also yield a reasonably good reconstruction, without having to calculate the optimal solutions numerically.
The performance of both the mirror filter and the set of optimal solutions (for different choices of $P^\prime$) can be evaluated straightforwardly using Eq. \eqref{eqn:snr_short}.
However, before comparing the results for various filters, there is another source of reconstruction error, specific to the MWA implementation of the PFB, that must first be considered.

@@

@@

We have implemented the synthesis filter defined in Eq. \eqref{eqn:synthesisfilter} as an optional component of the MWA-VCS beamforming software, \vcstools{}\footnote{\url{https://github.com/CIRA-Pulsars-and-Transients-Group/vcstools}}, described in \PaperI{}.
In this mode, a time series reconstructed from the beamformed fine channelised data is written out for each coarse channel as two separate polarisation streams (i.e. without converting to Stokes parameters) in the VDIF file format \citep{Whitney2009}.
This format was chosen because it is supported by the coherent de-dispersion functionality of DSPSR\footnote{\url{http://dspsr.sourceforge.net/}} \citep{VanStraten2011b}, a software suite for processing pulsar time series.

@@

\section{The FPGA implementation of the fine PFB}

>>>>

One of the great advantages of the critically sampled PFB is the existence of a mathematically equivalent but computationally efficient implementation.
It can be shown that Eq. \eqref{eqn:wdft} is equivalent to first segmenting the windowed time series into taps, summing their respective samples element-wise, and performing a single DFT on the resulting array (now also of size $K$), i.e.
\begin{equation}
    X_k[m] = \sum_{n=0}^{K-1} b_m[n]\,e^{-2\pi jkn/K},
    \label{eqn:pfb}
\end{equation}
where
\begin{equation*}
    b_m[n] = \sum_{\rho=0}^{P-1} h[K\rho - n]\,x[n + mM - K\rho].
\end{equation*}
A short proof of this equivalence is given in \citet{Harris2011}.
The procedure described by Eq. \eqref{eqn:pfb} is called the \textit{weighted overlap-add} algorithm, and is illustrated in Fig. \ref{fig:pfb}.
\begin{figure}[p]
    \centering
    (a) \\
    \includegraphics[width=\columnwidth]{cartoon-top.eps} \\
    (b) \\
    \includegraphics[width=\columnwidth]{cartoon.eps}
    \caption{A diagrammatic representation of the weighted overlap-add algorithm, as defined in Eq. \eqref{eqn:pfb}. Panel (a) shows the filter window being translated along a discretely sampled signal (in this case, containing a sinusoid and noise) with a step size of one tap. At each step, panel (b) shows how the signal (first row) is multiplied by a filter (second and third rows), and each tap is summed (bottom left) and Fourier transformed to produce the final spectrum (bottom right).}
    \label{fig:pfb}
\end{figure}

@@

\subsection{MWA implementation of the fine PFB}
\label{sec:mwa_pfb}

@@

In the context of an astrophysical signal, it is desirable to quantify the reconstruction error in terms of the signal-to-noise ratio (S/N).
If we assume that individual samples follow, and are dominated by, the same Gaussian noise statistics, then the reduction in S/N of a reconstructed sample can be estimated by considering the fraction of the power of the reconstructed sample that came from the original sample:
\begin{equation}
    \frac{[\text{S/N}]_\text{recon}}{[\text{S/N}]_\text{orig}}
      = \frac{\left(\sum\limits_{m = -\infty}^{\infty} f_m[n]\,h_m[-n]\right)^2}{
            \sum\limits_{s = -\infty}^{\infty}
            \left(\sum\limits_{m = -\infty}^{\infty} f_m[n]\,h_{m-s}[-n]\right)^2
        }.
    \label{eqn:snr_long}
\end{equation}
This expression is invariant under the substitution $n \rightarrow n + \lambda M$, $\lambda \in \mathbb{Z}$, which indicates that the (average) reconstruction error is only a function of where the sample in question falls within a tap.
Consequently, any synthesis filter (except the exact inverse of the analysis filter, if it exists) will introduce a ``ringing'' effect into the reconstructed time series that has a period equal to the size of the PFB tap.

@@

Minimising the loss of S/N is equivalent to solving Eq. \eqref{eqn:invertibility_condition} using least squares regression.
Since any given reconstructed sample only receives contributions from samples spaced one tap apart, Eq. \eqref{eqn:invertibility_condition} can be thought of as $M$ independent conditions, one for each tap position, $n = 0, 1, 2, \dots, M-1$.
Thus, considering each tap position separately, each condition can be expressed as a minimal matrix equation,
\begin{equation}
    H^{(n)}F^{(n)} = D,
\end{equation}
where $H^{(n)}$, $F^{(n)}$, and $D$ are matrices whose elements are given by
\begin{equation}
    \begin{aligned}
        H^{(n)}_{ij} &= h_{P-1+j-i}[-n], \\
        F^{(n)}_j    &= f_j[n], \\
        D_i          &= \delta^{(P^\prime+P)/2}_i,
    \end{aligned}
\end{equation}
where $P$ is the number of taps in the analysis filter, and the size of $F^{(n)}$ is set to the desired number of (non-zero) taps in the synthesis filter, $P^\prime$.
The indices are chosen such that, owing to the finite size of the analysis filter, the smallest $H^{(n)}$ that captures every non-trivial term in Eq. \eqref{eqn:invertibility_condition} is the $(P^\prime+P-1) \times P^\prime$ matrix
\begin{equation}
    H^{(n)} =
    \begin{bmatrix}
       h_{P-1}[-n] & 0 & \cdots & 0 \\
       h_{P-2}[-n] & h_{P-1}[-n] & \cdots & 0 \\
       \vdots & \ddots & \ddots & \vdots \\
       0 & \cdots & h_0[-n] & h_1[-n] \\
       0 & \cdots & 0 & h_0[-n]
    \end{bmatrix}
\end{equation}
with row number $i = (P^\prime + P)/2$ (in $H^{(n)}$ and $D$) corresponding to the $s = 0$ term\footnote{Note that since $P = 12$ for the PFB considered here, only the case where $P^\prime$ is even has been considered.}.

Once the matrices $H^{(n)}$, $F^{(n)}$, and $D$ have been defined, solution by least squares regression can proceed in the usual way, yielding the best-fit filter coefficients
\begin{equation}
    \hat{F}^{(n)} = \left({H^{(n)}}^T H^{(n)}\right)^{-1} {H^{(n)}}^T D,
\end{equation}
where ${H^{(n)}}^T$ is the transpose of $H^{(n)}$.
With this notation, the reconstruction error can be more concisely expressed
\begin{equation}
    \frac{[\text{S/N}]_\text{recon}}{[\text{S/N}]_\text{orig}}
      = \frac{(\hat{D}_{P-1})^2}{\hat{D}^T \hat{D}},
    \label{eqn:snr_short}
\end{equation}
where $\hat{D} = H^{(n)} \hat{F}^{(n)}$.

Fig. \ref{fig:synthfilters} shows the solutions found for the MWA's analysis filter defined in Eq. \eqref{eqn:filter}, for 12-, 18-, and 24-tap synthesis filter sizes.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{example-image-a}
    \caption{Least-squares solutions for different filter sizes \dots}
    \label{fig:synthfilters}
\end{figure}
Despite the complexity of the synthesis filters, they all share the same basic form as the analysis filter.
This suggests that choosing a synthesis filter that is the mirror image of the analysis filter (i.e. $f[n] = h[-n]$), might also yield a reasonably good reconstruction, without having to calculate the optimal solutions numerically.
The performance of both the (sub-optimal) mirror filter and the (optimal) set of least-squares solutions can be evaluated straightforwardly using Eq. \eqref{eqn:snr_short}.
However, before comparing the results for various filters, there is another source of reconstruction error, specific to the MWA implementation of the PFB, that must first be considered.

@@

The MWA implementation of the fine PFB differs from Eq. \eqref{eqn:pfb} in a few important respects, causing a reduction in the reconstructed S/N that would be present even if the synthesis filter perfectly inverted the analysis filter.
The most significant difference is that both the coefficients $b[n]$ and the final sum undergo a rounding operation, resulting in the approximate (i.e. less precise and biased) spectrum
\begin{equation}
    \tilde{X}_k[m] = \left\lfloor\sum_{n=0}^{K-1} \bigg\lfloor b_m[n] \bigg\rfloor_{\text{asym}} e^{-2\pi jkn/K}\right\rfloor,
\end{equation}
where $\lfloor\cdot\rfloor$ denotes rounding to the nearest integer, and $\lfloor\cdot\rfloor_{\text{asym}}$ denotes the asymmetrical rounding scheme described in detail in Appendix \ref{sec:fpga_pfb}.
Under these circumstances, Eq. \eqref{eqn:snr_long} cannot be used in its current form to estimate the signal loss, because it does not follow from the adjusted synthesis operation
\begin{equation}
    \hat{x}[n] = \sum_{m = -\infty}^{\infty} f[n - mM]\,
        \frac{1}{K} \sum_{k=0}^{K-1} \tilde{X}_k[m]\,e^{2\pi jkn/K},
    \label{eqn:ipfb_adjusted}
\end{equation}
which differs from Eq. \eqref{eqn:ipfb} only by the substitution $X_k[m] \rightarrow \tilde{X}_k[m]$.

@@

We have implemented the ``mirror'' synthesis filter as an optional component of the MWA-VCS beamforming software, \vcstools{}\footnote{\url{https://github.com/CIRA-Pulsars-and-Transients-Group/vcstools}}, described in \PaperI{}.
In this mode, a time series reconstructed from the beamformed fine channelised data is written out for each coarse channel as two separate polarisation streams (i.e. without converting to Stokes parameters) in the VDIF file format \citep{Whitney2009}.
This format was chosen because it is supported by the coherent de-dispersion functionality of DSPSR\footnote{\url{http://dspsr.sourceforge.net/}} \citep{VanStraten2011b}, a software suite for processing pulsar time series.

@@

\section{The FPGA implementation of the fine PFB}
\label{sec:fpga_pfb}
